---
title: "Partisan Dating Preferences Experiment"
output: html_notebook
---

This project is an experiment I ran in grad school. I was interested in how 
partisanship impacted dating preferences. In order to study this I set up a 
survey experiment. In the experiment I had participants view 9 dating profiles.
The information on the dating profile, including the partisanship, was randomized
for each participant. The respondent indicated on a scale of 1-7 how much they
would be interested in dating this individual. I recorded political info for the
respondent as well. I predcited that individuals who were partisan (Republicans
or Democrats) would be less interested in profiles with opposing partisanship. However
I did not think finding a member of the same party would be much of a boost. The
profiles were randomly assigned to be Republican, Democrat, or blank. The blank
served as the control in the experiment.

```{r}
rm(list = ls())

library(dplyr)

setwd("C:/Users/cayde/OneDrive/Data Science/R Projects/Dating Preferences")

dating_data <- read.csv("Dating Preferences Data.csv")
```

Because I am interested in how the political beliefs of the respondents effects
their dating preferences I drop all respondents who did not provide their political
information. 

```{r}

dating_data <- filter(dating_data, !is.na(Ideology) & !is.na(Party.ID)) #removing non response for party and ideology
dating_data <- dating_data[,-c(1:10)] #removing metadata
```

Here I get descriptive statistics of data I am interested in. Given that the profiles
had individuals of a certain age I restricted my survey to 18-25 year olds.
Party ID and Ideology are on a scale of 1-7 with 1 being Democrat and liberal and
7 being Republican and conservative. The mean is 4. Because the mean of my sample
for Party ID and ideology is below 4 I know that my sample skews liberal. This
is expected as the sample was taken from Mturk, which is known to be younger, 
more educated, and typically more liberal. The feeling thermometers of 63 vs. 44
confirm that individuals in my sample feel more warmly to Democrats than Republicans.

```{r}

#getting descriptive means
colMeans(dating_data[,names(dating_data) %in% c("Age", "Ideology", "Sex", "Party.ID", "Republican.Therm", "Democrat.Therm")])

```

In the data each profile has it's own variable name. I have 9 profiles of men
and 9 profiles of women. Each profile has an assigned partisanship. The profiles
are have names X1 through X9. F stands for female and M stands for Male. R stands
for Republican, D stands for Democrat, and C stands for control. The variable
X1FD_1 would indicate that the first female profile was assigned democrat. X6MC_1
indicates that the 6th male profile was assigned control. This naming scheme
allows me to know exactly which profiles each respondent saw.

In the code below I create an average rating for the profiles by their partisanship. 

```{r}

#Creating variables that are averages of respondents rating for Democrat, Republican and Control

dating_data$democrat_avg <- rowMeans(subset(dating_data, select = c(X1FD_1,X2FD_1,X3FD_1,X4FD_1,X5FD_1,X6FD_1,X7FD_1,X8FD_1,X9FD_1,X1MD_1,X2MD_1,X3MD_1,X4MD_1,X5MD_1,X6MD_1,X7MD_1,X8MD_1,X9MD_1)), na.rm = TRUE)

dating_data$republican_avg <- rowMeans(subset(dating_data, select = c(X1FR_1,X2FR_1,X3FR_1,X4FR_1,X5FR_1,X6FR_1,X7FR_1,X8FR_1,X9FR_1,X1MR_1,X2MR_1,X3MR_1,X4MR_1,X5MR_1,X6MR_1,X7MR_1,X8MR_1,X9MR_1)), na.rm = TRUE)

dating_data$control_avg <- rowMeans(subset(dating_data, select = c(X1FC_1,X2FC_1,X3FC_1,X4FC_1,X5FC_1,X6FC_1,X7FC_1,X8FC_1,X9FC_1,X1MC_1,X2MC_1,X3MC_1,X4MC_1,X5MC_1,X6MC_1,X7MC_1,X8MC_1,X9MC_1)), na.rm = TRUE)

```

I would like to remove the profiles after I have aggregated the data so that I
will have a clean data set. The profiles all begin with X. Here I rename some
variables that have an X in them so I can easily identify the profile vars with 
the letter X.

```{r}

# All profile variables start with "X"
# Renaming other variables that have an "X" in them to isolate the profiles
dating_data <- rename(dating_data, "Orientation" = "Sexual.Orientation")
dating_data <- rename(dating_data, "Gender" = "Sex")

```

Before I remove the profiles I want to aggregate them again. But rather than 
having a continuous measure I would like to convert it to a binary one (interested
vs. not interested). Many dating apps have a simple binary feature (left swipe vs.
right swipe). This just gives me another way of reporting the data. 

```{r}

profile_vars <- names(select(dating_data, contains("X")))

# Converting Likert scale to binary variable

for(column in profile_vars){
  dating_data[,column] <- recode(dating_data[,column], `7` = 1,
                               `6` = 1,
                               `5` = 1,
                               `4` = 1,
                               `3` = 0,
                               `2` = 0,
                               `1` = 0)
}

#Creating variables that are averages of respondents rating for Democrat, Republican and Control

dating_data$democrat_interest <- rowMeans(subset(dating_data, select = c(X1FD_1,X2FD_1,X3FD_1,X4FD_1,X5FD_1,X6FD_1,X7FD_1,X8FD_1,X9FD_1,X1MD_1,X2MD_1,X3MD_1,X4MD_1,X5MD_1,X6MD_1,X7MD_1,X8MD_1,X9MD_1)), na.rm = TRUE)
dating_data$republican_interest <- rowMeans(subset(dating_data, select = c(X1FR_1,X2FR_1,X3FR_1,X4FR_1,X5FR_1,X6FR_1,X7FR_1,X8FR_1,X9FR_1,X1MR_1,X2MR_1,X3MR_1,X4MR_1,X5MR_1,X6MR_1,X7MR_1,X8MR_1,X9MR_1)), na.rm = TRUE)
dating_data$control_interest <- rowMeans(subset(dating_data, select = c(X1FC_1,X2FC_1,X3FC_1,X4FC_1,X5FC_1,X6FC_1,X7FC_1,X8FC_1,X9FC_1,X1MC_1,X2MC_1,X3MC_1,X4MC_1,X5MC_1,X6MC_1,X7MC_1,X8MC_1,X9MC_1)), na.rm = TRUE)

```

After recoding the profiles and aggregating them to a variable called 'interest'
I now remove all the profiles as well as any observations with missing data.

```{r}

# Removing individual profiles

dating_data <- select(dating_data, !contains("X"))

dating_data <- na.omit(dating_data)
```

I also need to convert the 7 pt Party ID to 
the typical Republican, Democrat, and Independent. This makes the results of the
experiment more easy to report. 

```{r}
#converting 7 point PID to party

dating_data$Party <- NA
dating_data$Party[dating_data$Party.ID < 4] <- "Democrat"
dating_data$Party[dating_data$Party.ID == 4] <- "Independent"
dating_data$Party[dating_data$Party.ID > 4] <- "Republican"

```

The final transformation I need to make to the data is to convert it into 'long'
format. Currently it is in a wide format. Each respondent has an average rating
for Democrat profiles, Republican profiles, and control profiles. These represent
3 different columns. In order to do an ANOVA, I need to put the data into long format.
One column will indicate the partisanship of the profile and another column will
have the average rating. There will be three rows per respondent. In this manner
I can treat the rating of the profile as a dependent variable and the partisanship
of the profile as an independent variable. 



```{r}

# AOV

# Converting data to 'long' format
dating_data_long <- gather(dating_data, key = "partisanship_of_profile", value = "average_rating", democrat_avg:control_avg)

```

Before doing the analysis let's visualize the data

```{r}
ggplot(dating_data_long, aes(y=average_rating, x=Party, fill = partisanship_of_profile)) +
  geom_boxplot() +
  labs(y="Rating") +
  scale_fill_manual(name = "Partisanship of Profile", 
                    labels = c("Control", "Democrat", "Republican"),
                    values = c("gray", "blue", "red"))

```

It appears as though Democrats are significantly less interested in dating 
Republicans and vice versa. Independents on the other hand do no have a preference.
Interestingly the differences between the control and the in-party profiles show
little difference for both Democrats and Republicans. Let's perform statistical 
analysis to see if these effects are significant. 

Now I move on to the ANOVA. I could simply put partisanship as a control variable,
but I am interested in seeing the differences between Republicans and Democrats.
Perhaps one party is more concerned with whom they data than the other party. 
I run an ANOVA regressing the rating on partisanship of the profile. Remember their
are three types of profiles (Democrats, Republicans, and control), thus I need an
ANOVA rather than the t-test.

Looking at the Democratic respondents we can see that the partisanship of the profile
was a significant predictor of their rating. Performing Tukey HSD tests we see
where the differences lie. As predicted Republican profiles were viewed much less
favorably compare to control and Democratic profiles. However, Democrats did not
prefer Democrat profiles compared to control profiles. 

```{r}

dem_pref <- aov(lm(average_rating ~ partisanship_of_profile, data = dating_data_long[dating_data_long$Party == "Democrat",]))
summary(dem_pref)
TukeyHSD(dem_pref)

```

I do not have any expect and strong effects for independents. While it is possible
that independents dislike both Republican and Democratic profiles this would make
dating very difficult for them. Indeed, I find no significant effects for partisanship
and the TukeyHSD tests show that there are no large differences between the groups.

```{r}

ind_pref <- aov(lm(average_rating ~ partisanship_of_profile, data = dating_data_long[dating_data_long$Party == "Independent",]))
summary(ind_pref)
TukeyHSD(ind_pref)

```

Finally I look at Republicans. The partisanship of the profile did have a significant
effect. Looking at the TukeyHSD tests we see a similar story to the Democrats:
Republicans don't particularly like Republican profiles more than control profiles,
however Republicans like Democrat profiles significantly less than both Republican
and Democrat profiles. The p-value of the Republican and Democrat profile is above
the conventional 5% but just barely. Considering that I only had 21 republicans
I am confident that this is due to small sample size. Interestingly the Republicans
actually liked the control profile more than the Repbulican profile. This difference,
however, is extremely small an nowhere near statisticaly significant. This observation
is also consistent with my expectation that being of the same party does not necessarily
make a profile more attractive. Instead, profiles of the opposing party are 
liked significantly less. 

```{r}

rep_pref <- aov(lm(average_rating ~ partisanship_of_profile, data = dating_data_long[dating_data_long$Party == "Republican",]))
summary(rep_pref)
TukeyHSD(rep_pref)

```

In summation this data shows that partisans are much less willing to data members
of the opposite party, but that they do no necessarily prefer members of their 
own party. This is significant because it aligns with what Political Scientists
calle "Negative Partisanship", the idea that dislike of the opposing party is more
consequential than like for ones own. The experimetal nature of this research 
ensures that it was in fact the partisanship that was influencing dating preferences.
Some scholars have suggested that couples are politically similar due to convergence 
(they adopt each others political beliefs over time), homophily (one's dating pool
is usually demographically similar to oneself) and correlation (individuals are
selecting on other factors such as race or religion, which are simply correlated
with politics). My study shows that selection may be playing a role. That is 
individuals specificially avoid dating someone with opposing political beliefs. 