---
title: "Scraping Texas Election Data"
output: html_notebook
---

In this project I am interested in scraping data for Texas Elections. Each county
in Texas hosts elections every two years for various offices such as Commissioner,
Judge, District Attorney etc. Going to this website: 
https://imis.county.org/imis/TACMember/Elections/General_Election_Results.aspx
you can enter the year and the county you are interested and click "Refresh."
The results will give you a table of the 1) position title 2) all of the eligible
candidates 3) their respective party 4) voter number and 5) vote percentage. 

For a human to navigate to a county of interest is relatively simple; you simply
select the year and county from a drop-down list and hit refresh. However, if you
wanted to automate the process you would need to write code that can select the
county you want, 'hit' the refresh button, collect the results, and then repeat
the process for every other county. 

Using rvest package I have written code to accomplish this. I even have code for
when their is missing data for the counties. 

```{r}
rm(list = ls())

setwd("C:/Users/cayde/OneDrive/Graduate Studies/Misc/RM/Marshall")

library(tidyverse)
library(rvest)

```

The first thing to do is set the url and session. This is standard practice with
Rvest. The url can be set to a string of the URL. You then create a session variable
by using the html_session function and put the url inside it as one of the arguments.
Normally this would suffice if you were just interested in scraping data from a 
static website. However, as explained earlier, we need to choose values and 
refresh the page. For this we will need to create a form. Here I have created a 
variable named 'form.' by putting my session variable inside the html_form function.

```{r}

#Also be sure to rename your variables to the year when creating an empty dataframe, putting data in that data frame "new_year[i,column] <- results...."
## Viewing the data frame "view(new_year)" and saving "write.csv <- (new_year, ..."

url <- "https://imis.county.org/imis/TACMember/Elections/General_Election_Results.aspx"
session <- html_session(url)
form <- html_form(session)[[1]]

```

Now that the form is created I can access different elements of the page using the
form. The syntax is a little complicated, but you can arrive at what I need through
trial and error. To do this I simply go to the form object in my Global Environment 
in R. After clicking on it I can manually navigate to see where what I want is
located.

In this case I need a list of counties to loop through. Using the navigation 
technique I find that it is located under options, which is in the 30th' row 
under the fields value. I can access it with the following code:

```{r}
counties <- names(form$fields[[30]]$options)

```

Now that I have a list of counties a create a dataframe to put my data in. I
then create a loop that loops through the counties. Inside the loop I use the 
set_values function. The first argument it takes is the form, which I conveniently
named form. The remaining arguments are looking to set the fields in the form. 
This is the equivalent of manually choosing the year and county from a drop-down
list. But of course here we choose these by code. Using the navigation technique
I find the name of the field and set it equal to the appropriate value. For 
whatever reason the value is not the name of the county, but instead each county
has a number assigned to it that goes in increments of 10. The first county 'Anderson'
has a value of 10, the second county 'Andrews' has a value of 20, and this 
continues until the last county Zavala that has a value of 2540. Taking advantage
of this pattern I simply put loop index and multiply it by 10. The second field
is simply the year and I set it to the year I want. 

After I have chosen the values I need to do the code equivalent of hitting the 
refresh button. I do this with the submit_form argument. It takes the argument of
the session as well as the form that I have just filled out. It's final argument
is the value of the refresh button which includes "SubmitButton" in it's name. 

After the form is submitted the results appear. Next I record the results
with the html_table function. The html table does not put the data in a dataframe
but rather a list. My next step is to put the results in a dataframe, which I do
in the for(webpage in 1:length....) loop. Before I do I check to see if the county
is missing from the webpage. Some of the results have missing counties. If this 
is the case you can put the numeric value of the county in the missing county 
vector and when it comes time to get the information for the html_table, it will
simply skip this loop if 

Navigating through the html table with the results I find that all the information is
in the X1, X2, X3, and X4 variables in the third row. These are indexed using 
results[[3]][[X1]], results[[3]][[X2]] and so on. I create a vector for X1, X2, 
X3, and X4 and then loop through these variable names to retrieve the information. 
In the loop I assign the information to a column. The column assignment is a function 
of the county and the variable name. For example, X1, X2, X3, and X4 will be assigned 
to columns 1-4 for the first county. X1, X2, X3, and X4 will be assigned to columns 
5-8 for the second county and so on. After this is done I print the name of the 
county, this is primarily so I can see that my loop is running. After this the 
next county is chosen, a new form is created, and the process is repeated. 

```{r}
general_2020 <- data.frame()

missing_counties <- c()

for(county in 1:length(counties)){
  
  form <- set_values(form, "ctl01$TemplateBody$WebPartManager1$gwpciNewReportDisplayCommon$ciNewReportDisplayCommon$ParamSheet1$Input1$ctl00" = county*10,
                     "ctl01$TemplateBody$WebPartManager1$gwpciNewReportDisplayCommon$ciNewReportDisplayCommon$ParamSheet1$Input0$ctl00" = 2020)
  
  session <- submit_form(session = session, form = form, submit = "ctl01$TemplateBody$WebPartManager1$gwpciNewReportDisplayCommon$ciNewReportDisplayCommon$ParamSheet1$SubmitButton")
  
  results <- html_table(session, fill = TRUE)
  
  column_webpage <- c("X1", "X2", "X3", "X4")
  webpage <- 1
  
  for(webpage in 1:length(column_webpage)){
    column <- county*4 - (4 - webpage)
    general_2020[1:length(results[[3]][[column_webpage[webpage]]]),column] <- results[[3]][[column_webpage[webpage]]]
  }
  print(counties[county])
}

view(general_2020)

write.csv(general_2020, file = "Texas General 2020.csv")

```

The result of this is all of the results from all the counties for a given election
year. I write the file to a desired location on my computer. 



